Factors Influencing Abundance & Virus Isolation in Mosquitoes

### 2024-04-25 R/analysis-code/analysis.qmd
further edits and file cleaning

```{r}
# Load the PCR data set which includes information on mosquito pools tested for viruses, 
# results of those tests, and the date of collection.
pcr <- read_csv(here("data", "processed-data", "pcr_data_expanded_filtered.csv"))

@@

weather_weekly <- wx_selected %>%
  arrange(week_num) %>%
  mutate(
    Lagged_Avg_TMAX = lag(Avg_TMAX, 1),
    Lagged_Total_PRCP = lag(Total_PRCP, 1),
    Binary_PRCP = if_else(Lagged_Total_PRCP > 0, 1, 0)
  )
# Aggregate the PCR data by week to determine if there was at least one positive result in each week.
# This transformation reduces the detailed test results into a weekly summary,
# indicating the presence or absence of positive tests each week.
pcr_weekly <- pcr %>%
  group_by(week_num) %>%
  summarise(Positive_Result = max(as.integer(Is_Positive), na.rm = TRUE))

@@

```{r}
pcr <- read_csv(here("data", "processed-data", "pcr_data_expanded_filtered.csv"))
wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))

@@

# Load and prepare mosquito data
first_positive_dates <- pcr_data_expanded_filtered %>%
  filter(Result == "Positive") %>%
  group_by(Year = year(`Collection Date`)) %>%
  summarise(First_Positive_Date = min(`Collection Date`), .groups = 'drop')

@@

# # load the dataset with a unique dataframe name
# mir_2part_all <- read_csv(here("data", "processed-data", "all_mir_clean_joined.csv"))

>>>>

```{r}
# Load the PCR data set which includes information on mosquito pools tested for viruses, 
# results of those tests, and the date of collection.
pcr <- read_csv(here("data", "processed-data", "filtered_pcr_data.csv"))

@@

weather_weekly <- wx_selected %>%
  arrange(week_num) %>%
  mutate(
    Lagged_Avg_TMAX = lag(Avg_TMAX, 1),
    Lagged_Total_PRCP = lag(Total_PRCP, 1),
    Binary_PRCP = if_else(Lagged_Total_PRCP > 0, 1, 0)
  )
# Aggregate the PCR data by week to determine if there was at least one positive result in each week.
# This transformation reduces the detailed test results into a weekly summary,
# indicating the presence or absence of positive tests each week.
pcr_weekly <- pcr %>%
  group_by(week_num) %>%
  summarise(Positive_Result = ifelse(all(is.na(Is_Positive)), 0, max(as.integer(Is_Positive), na.rm = TRUE)))

@@

```{r}
pcr <- read_csv(here("data", "processed-data", "pcr_data.csv"))
wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))

@@

# Load and prepare mosquito data
first_positive_dates <- pcr %>%
  filter(Result == "Positive") %>%
  group_by(Year = year(`Collection Date`)) %>%
  summarise(First_Positive_Date = min(`Collection Date`), .groups = 'drop')

@@

# load the dataset with a unique dataframe name
mir_2part_all <- read_csv(here("data", "processed-data", "all_mir_clean_joined.csv"))

### 2024-04-25 R/eda-code/EDA_unified.qmd
further edits and file cleaning

@@

# Save as RDS
saveRDS(MIR_by_week, here("data", "processed-data", "rds", "MIR_by_week.rds"))
View(MIR_by_week)

@@

# Check for overdispersion
if(vmr > 1) {
  cat("Evidence of overdispersion: VMR is greater than 1.\n")
} else {
  cat("No evidence of overdispersion: VMR is not greater than 1.\n")
}
```
#### Given the excess zeros and the overdispersion in the MIR data, it may be appropriate to consider a Generalized Additive Model (GAM). GAM allow for 2-part models that can models 
# Predictors
#### Closer look at the possible predictors for the model
```{r}
# Load the dataset:
# combined_data_mir_all: Contains weekly MIR data and possible predictors for the model.
joined_mir <- read_csv(here("data", "processed-data", "joined_data_mir_co2.csv"))

# Assuming 'joined_mir' is your dataframe
continuous_vars <- joined_mir[, c("Avg_TMAX", "Total_PRCP", "Hrs_Above_80_lag", 
                                  "Hrs_Above_75_lag", "Hrs_Below_50_lag", 
                                  "Avg_HourlyDryBulbTemp_lag", "Avg_HourlyRelativeHumidity_current", 
                                  "Avg_HourlyWetBulbTemperature_current", "Avg_HourlyWindSpeed_current", 
                                  "DSCI_fema", "average_daylight_hours", 
                                  "Average_MEL", "Total_MEL", "MEL_Prop")]

>>>>

#save as csv
write_csv(filtered_pcr_data, here("data", "processed-data", "filtered_pcr_data.csv"))

@@

# Save as RDS
saveRDS(MIR_by_week, here("data", "processed-data", "rds", "MIR_by_week.rds"))
str(MIR_by_week)

@@

# Check for overdispersion
if(vmr > 1) {
  cat("Evidence of overdispersion: VMR is greater than 1.\n")
} else {
  cat("No evidence of overdispersion: VMR is not greater than 1.\n")
}
```
#### Given the excess zeros and the overdispersion in the MIR data, it may be appropriate to consider a Generalized Additive Model (GAM). GAM allow for 2-part models that can models 
# Predictors
#### Closer look at the possible predictors for the model
```{r}
# Load the dataset:
# combined_data_mir_all: Contains weekly MIR data and possible predictors for the model.
joined_mir <- read_csv(here("data", "processed-data", "joined_data_mir_co2.csv"))
joined_mir
# Assuming 'joined_mir' is your dataframe
continuous_vars <- joined_mir[, c("Avg_TMAX", "Total_PRCP", "Hrs_Above_80_lag", 
                                  "Hrs_Above_75_lag", "Hrs_Below_50_lag", 
                                  "Avg_HourlyDryBulbTemp_lag", "Avg_HourlyRelativeHumidity_current", 
                                  "Avg_HourlyWetBulbTemperature_current", "Avg_HourlyWindSpeed_current", 
                                  "DSCI_fema", "average_daylight_hours", 
                                  "Average_MEL", "Total_MEL", "MEL_Prop")]

# Summarize missing values for each variable
missing_values <- sapply(joined_mir[, c("Avg_TMAX", "Total_PRCP", "Hrs_Above_80_lag", 
                                        "Hrs_Above_75_lag", "Hrs_Below_50_lag", 
                                        "Avg_HourlyDryBulbTemp_lag", "Avg_HourlyRelativeHumidity_current", 
                                        "Avg_HourlyWetBulbTemperature_current", "Avg_HourlyWindSpeed_current", 
                                        "DSCI_fema", "average_daylight_hours", 
                                        "Average_MEL", "Total_MEL", "MEL_Prop")], function(x) sum(is.na(x)))
print(missing_values)

### 2024-04-25 R/processing-code/MADAproject_pt2_ruiz.qmd
further edits and file cleaning

# Loading data
## Mosquito trap data
### This dataset contains trap counts by species for all mosquito traps set in the Bristol County from 2007-2023 this dataset will undergo extensive processing below.
#### The dataset was obtained though and open records request with the Commonwealth of MA.
```{r trap-data-import}
# Construct the path to the Excel file
#file_path_mos <- here("data", "raw-data", "MOSQ_MADA.xlsx")
file_path_mos <- here("data", "raw-data", "MOSQ_MADA.xlsx")
file_exists_check <- file.exists(file_path_mos)
file_path_mos <- here("data", "raw-data", "MOSQ_MADA.xlsx")

@@

# Save as RDS
saveRDS(pcr_data_expanded, here("data", "processed-data", "rds", "pcr_data"))

# Save as CSV
write_csv(pcr_data_expanded, here("data", "processed-data", "pcr_data.csv"))

@@

## Mosquito trap sites
### Location for the mosquito trap sites
```{r}
# Construct the path to the file
file_path_trap = here("data", "processed-data", "BCMCP_trap_sites_2021.xlsx")

@@

# Ensure hourly_wx is read if not already done
hourly_wx <- read_csv(hourly_wx_path, show_col_types = FALSE)
View(hourly_wx)
hourly_wx_clean <- hourly_wx %>%
  mutate(Year = year(DATE),
         ISO_Week = isoweek(DATE),
         week_num = sprintf("%d-%02d", Year, ISO_Week)) %>%
  filter(REPORT_TYPE...3 != "SOD", REPORT_TYPE...3 != "SOM", REPORT_TYPE...3 != "AUTO") %>%
  select(STATION, DATE, HourlyDryBulbTemperature, HourlyRelativeHumidity, HourlyWetBulbTemperature, HourlyWindSpeed, week_num)

str(hourly_wx_clean)

@@

# read file with drought data by county for FEMA Region 1-4 for the years 2007-2023
drought_fema1_4 = read_csv(here("data", "raw-data", "comp_stat_fema1_4.csv"), show_col_types = FALSE)
#View(drought_fema1_4)
# Add week number based on MapDate for both files
# This code ensures that the ISO 8601 standard is applied to the week number calculation
# Week 1 of any year is the week that contains January 4th, or equivalently, it's the week that contains the first Thursday of January.
# Weeks start on Monday and end on Sunday.
# The last week of the year, week 52 or 53, is the one that contains December 28th.
# we will use week_num to join tables
# the format will be YYYY-WW. Week 8 of 2019 will look like: 2019-08

@@

drought_ma_counties <- drought_ma_counties %>%
  mutate(
    # Convert MapDate from a numeric format to a Date object
    MapDate = as.Date(as.character(MapDate), format = "%Y%m%d"),
    # Create a week_num column formatted as "YYYY-WW"
    week_num = paste0(year(MapDate), "-", sprintf("%02d", isoweek(MapDate))),
    # Calculate the DSCI, assuming D0 to D4 are numeric
    DSCI_ma = rowSums(select(., D0:D4), na.rm = TRUE)
  )
#View(drought_ma_counties)

@@

```{r}
# Construct the path to the human case data file
file_path_human <- here("data", "processed-data", "human_cases.xlsx")

@@

## Bird data
#### This dataset comes from the annual North American Breeding Bird Survey that is conducted in a joint effort by: USGS Eastern Ecological Science Center, Environment and Climate Change Canada, and the Mexican National Commission for the Knowledge and Use of Biodiversity.
#### There are 2 (5 and 11) routes in Bristol County, MA. The data is from 2007-2022. However, route 11 is mission data for 2013 and route 5 ends in 2019.
#### For route 11, 2013, we will use the data from 2012 and 2014 to estimate the missing data.
for route 5, 2020-2022, we will use the data from 2019 to estimate the missing data.
```{r}
# Load the dataset
file_path_bird <- here("data", "raw-data", "Massach.csv")
bird_data_raw <- read.csv(file_path_bird)

# Filter and clean up the data
bird_bbs <- bird_data_raw %>%
  filter(Route %in% c(5, 11), Year >= 2006 & Year <= 2022) %>%
  select(-AOU, -Count10, -Count20, -Count30, -Count40, -Count50)

# Summarize the data
bird_bbs_summarized <- bird_bbs %>%
  group_by(Route, Year) %>%
  summarize(
    StopTotal = sum(StopTotal, na.rm = TRUE),
    SpeciesTotal = sum(SpeciesTotal, na.rm = TRUE),
    .groups = 'drop'
  )

# Check for missing data within Routes 5 and 11
missing_data <- map_df(c(5, 11), ~tibble(
  Route = .x,
  missing_years = setdiff(2006:2022, bird_bbs_summarized$Year[bird_bbs_summarized$Route == .x])
))

# Identify all combinations of Route and Year within the range
all_combinations <- expand.grid(Route = unique(bird_bbs_summarized$Route),
                                Year = 2006:2022)

# Join with the summarized data to identify missing rows
bird_bbs_full <- left_join(all_combinations, bird_bbs_summarized, by = c("Route", "Year"))

# Impute missing data by taking the average of the previous and next year for each route
bird_bbs_imputed <- bird_bbs_full %>%
  group_by(Route) %>%
  arrange(Route, Year) %>%
  mutate(
    StopTotal = ifelse(is.na(StopTotal), (lag(StopTotal) + lead(StopTotal)) / 2, StopTotal),
    SpeciesTotal = ifelse(is.na(SpeciesTotal), (lag(SpeciesTotal) + lead(SpeciesTotal)) / 2, SpeciesTotal)
  ) %>%
  ungroup()

# Handling the last records for Route 5 (2020-2022)
# Since there's no future data to average, you might replicate the most recent data (e.g., from 2019)
# Or consider another imputation method suitable for your analysis and data characteristics

# Find the most recent year's data for Route 5
recent_data_route_5 <- bird_bbs_imputed %>% 
  filter(Route == 5, Year == max(Year[Route == 5 & !is.na(StopTotal)]))

# Create a template for the missing years
missing_years_route_5 <- tibble(Route = 5, Year = 2020:2022)

# Replicate the most recent data for the missing years
imputed_missing_years_route_5 <- bind_rows(lapply(missing_years_route_5$Year, function(y) {
  mutate(recent_data_route_5, Year = y)
})) %>%
  bind_rows(bird_bbs_imputed) %>%
  arrange(Route, Year)

# Check if any adjustments or specific imputations are needed
print(imputed_missing_years_route_5)

bird_clean <- imputed_missing_years_route_5 %>%
  drop_na()

# Save the cleaned bird data as an csv
write_csv(bird_clean, here("data", "processed-data", "bird_clean.csv"))
```

@@

# Ensure there's no duplication in the conversion attempt
# Step 3: Convert list columns to numeric, replacing NULL or list elements without values with 0
# Utilizing sapply for a direct approach to handle potential list-columns
wide_data <- wide_data %>%
  mutate(across(.cols = all_of(species_columns), 
                .fns = ~ map_dbl(.x, function(item) {
                  # Handle NULL items directly
                  if (is.null(item)) {
                    return(0)
                  # Handle vectors: convert NA to 0, and take the first item if it's a vector
                  } else if (length(item) > 1 || is.na(item)) {
                    item[is.na(item)] <- 0
                    return(as.numeric(item[1]))
                  } else {
                    return(as.numeric(item))
                  }
                })))
# Step 4: Add week_num column
mosquito_wide <- wide_data %>%
  mutate(
    `collection date` = as.Date(`collection date`),
    week_num = paste0(year(`collection date`), "-", sprintf("%02d", isoweek(`collection date`)))
  )

# Output the transformed data to inspect its structure
head(mosquito_wide)

@@

# Save the wide-format data to the specified path
write_csv(wide_data, file_path_wide)

@@

# Save the wide-format data as an RDS file
saveRDS(wide_data, file_path_wide_rds)

# Now we will filter the the mosquito data to match the time period of the virus data. 
# Since mosquito populations can be affected by the previous year's populations, we will filter for 2007-2023

# Filter wide_data for the specified date range
filtered_mosquito_data <- wide_data %>%
  filter(`collection date` >= as.Date("2007-01-01") & `collection date` <= as.Date("2023-12-31"))

# Define the path for the CSV file
file_path_filtered_csv <- here("data", "processed-data", "mosquito_wide_filtered.csv")

# Save the filtered wide-format data to the specified CSV path
write_csv(filtered_mosquito_data, file_path_filtered_csv)
View(mosquito_wide_filtered.csv)
# Define the path for the RDS file
file_path_filtered_rds <- here("data", "processed-data", "rds", "mosquito_wide_filtered.rds")

# Save the filtered wide-format data as an RDS file
saveRDS(filtered_mosquito_data, file_path_filtered_rds)

@@

# Read the data
trap_proportion <- read_csv(trap_wide_path) %>%
  mutate(Year = year(as.Date(`collection date`, format = "%Y-%m-%d")))

# Assuming all species columns are next to each other, calculate the proportion of MEL
proportion_mel_by_year <- trap_proportion %>%
  # Calculate row-wise totals for all species
  rowwise() %>%
  mutate(Total_All = sum(c_across(ABS:OCH), na.rm = TRUE)) %>%
  ungroup() %>%
  # Then group by year and summarize
  group_by(Year) %>%
  summarise(Total_MEL = sum(MEL, na.rm = TRUE),
            Total_All = sum(Total_All, na.rm = TRUE), # Sum of all species by year
            MEL_Prop = Total_MEL / Total_All) # Proportion of MEL

@@

# Calculate the proportion of MEL by week
# Assuming all species columns are next to each other, calculate the proportion of MEL by week_num
proportion_mel_by_week <- trap_proportion %>%
  # Calculate row-wise totals for all species
  rowwise() %>%
  mutate(Total_All = sum(c_across(ABS:OCH), na.rm = TRUE)) %>%
  ungroup() %>%
  # Then group by week_num and summarize
  group_by(week_num) %>%
  summarise(Total_MEL = sum(MEL, na.rm = TRUE),
            Total_All = sum(Total_All, na.rm = TRUE), # Sum of all species by week_num
            MEL_Prop = Total_MEL / Total_All) # Proportion of MEL

@@

```{r}
# Load the datasets
yearly_mel_ave <- readRDS(here("data", "processed-data", "rds", "yearly_mel_ave.rds"))
summed_prcp_by_year_filtered <- readRDS(here("data", "processed-data", "rds", "summed_prcp_by_year_filtered_2013_2020.rds"))

@@

plot_mel_precip = ggplot(data = combined_data) +
  geom_col(aes(x = factor(Year), y = Average_MEL), fill = "blue") +
  geom_line(aes(x = factor(Year), y = Total_PRCP, group = 1), color = "red", linetype = "dashed") +
  geom_point(aes(x = factor(Year), y = Total_PRCP), color = "red", size = 3) +
  geom_hline(yintercept = 45.6, linetype = "dotted", color = "darkgreen", size = 1) +
  annotate("text", x = Inf, y = 45.6, label = "Avg Precip: 45.6 inches", 
           hjust = 3.2, vjust = -1.1, color = "darkgreen", size = 3.5, fontface = "bold", angle = 0) +
  scale_y_continuous(name = "Average Count of MEL", 
                     sec.axis = sec_axis(~ ., name = "Total Precipitation (inches)")) +
  labs(title = "Average Count of MEL Mosquitoes and Total Precipitation by Year",
       x = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.title.x = element_text(face = "bold"))
plot_mel_precip

@@

### Calcualte the average moon phase by week_num.
#### some resaecrh suggests that moon phase can affect mosquito activity.
##### The moon phase is calculated using the getMoonIllumination function from the lunar package. The function returns the percentage of the moon's illuminated surface area and the moon's phase angle. The phase angle is used to determine the moon phase, which can be one of the following: "New Moon", "First Quarter", "Full Moon", or "Last Quarter". The phase angle is converted to a value between 0 and 1, where 0 corresponds to a New Moon, 0.5 corresponds to a Full Moon, and 1 corresponds to a New Moon. The average moon phase for each week is calculated by taking the mean of the moon phase values for each day in the week.

```{r}
# Generate a sequence of dates from 2007-01-01 to 2023-12-31
dates <- seq(as.Date("2007-01-01"), as.Date("2023-12-31"), by = "day")

# Function to get moon phase for a given date
get_moon_phase <- function(date) {
  illumination <- getMoonIllumination(date)
  phase <- illumination$phase
  return(phase)
}

# Calculate moon phase for each date
moon_phases <- sapply(dates, get_moon_phase)

# Create a data frame with dates and their corresponding moon phases
moon_phases_df <- data.frame(Date = dates, MoonPhase = moon_phases)

# View the first few rows of the data frame
str(moon_phases_df)

# Add week_num column to moon_phases_df
moon_phases_df <- moon_phases_df %>%
  mutate(
    week_num = paste0(year(Date), "-", sprintf("%02d", isoweek(Date)))
  )

# Calculate average moon phase by week_num
average_moon_phase_by_week <- moon_phases_df %>%
  group_by(week_num) %>%
  summarise(AverageMoonPhase = mean(MoonPhase, na.rm = TRUE))

# View the results
head(average_moon_phase_by_week)

>>>>

# Loading data
## Mosquito trap data
### This dataset contains trap counts by species for all mosquito traps set in the Bristol County from 2007-2023 this dataset will undergo extensive processing below.
#### The dataset was obtained though and open records request with the Commonwealth of MA.
```{r trap-data-import}
# Construct the path to the Excel file
file_path_mos <- here("data", "raw-data", "MOSQ_MADA.xlsx")
file_exists_check <- file.exists(file_path_mos)
file_path_mos <- here("data", "raw-data", "MOSQ_MADA.xlsx")

@@

# Save as RDS
saveRDS(pcr_data, here("data", "processed-data", "rds", "pcr_data.rds"))

# Save as CSV
write_csv(pcr_data, here("data", "processed-data", "pcr_data.csv"))

@@

## Mosquito trap sites
### Location for the mosquito trap sites
```{r}
# Construct the path to the file
file_path_trap = here("data", "raw-data", "BCMCP_trap_sites_2021.xlsx")

@@

# Ensure hourly_wx is read if not already done
hourly_wx <- read_csv(hourly_wx_path, show_col_types = FALSE)
str(hourly_wx)
hourly_wx_clean <- hourly_wx %>%
  mutate(Year = year(DATE),
         ISO_Week = isoweek(DATE),
         week_num = sprintf("%d-%02d", Year, ISO_Week)) %>%
  filter(REPORT_TYPE...3 != "SOD", REPORT_TYPE...3 != "SOM", REPORT_TYPE...3 != "AUTO") %>%
  select(STATION, DATE, HourlyDryBulbTemperature, HourlyRelativeHumidity, HourlyWetBulbTemperature, HourlyWindSpeed, week_num)

@@

# read file with drought data by county for FEMA Region 1-4 for the years 2007-2023
drought_fema1_4 = read_csv(here("data", "raw-data", "comp_stat_fema1_4.csv"), show_col_types = FALSE)

# Add week number based on MapDate for both files
# This code ensures that the ISO 8601 standard is applied to the week number calculation
# Week 1 of any year is the week that contains January 4th, or equivalently, it's the week that contains the first Thursday of January.
# Weeks start on Monday and end on Sunday.
# The last week of the year, week 52 or 53, is the one that contains December 28th.
# we will use week_num to join tables
# the format will be YYYY-WW. Week 8 of 2019 will look like: 2019-08

@@

drought_ma_counties <- drought_ma_counties %>%
  mutate(
    # Convert MapDate from a numeric format to a Date object
    MapDate = as.Date(as.character(MapDate), format = "%Y%m%d"),
    # Create a week_num column formatted as "YYYY-WW"
    week_num = paste0(year(MapDate), "-", sprintf("%02d", isoweek(MapDate))),
    # Calculate the DSCI, assuming D0 to D4 are numeric
    DSCI_ma = rowSums(select(., D0:D4), na.rm = TRUE)
  )

@@

```{r}
# Construct the path to the human case data file
file_path_human <- here("data", "raw-data", "human_cases.xlsx")

@@

@@

# Ensure there's no duplication in the conversion attempt
# Step 3: Convert list columns to numeric, replacing NULL or list elements without values with 0
# Utilizing sapply for a direct approach to handle potential list-columns
wide_data <- wide_data %>%
  mutate(across(.cols = all_of(species_columns), 
                .fns = ~ map_dbl(.x, function(item) {
                  # Handle NULL items directly
                  if (is.null(item)) {
                    return(0)
                  # Handle vectors: convert NA to 0, and take the first item if it's a vector
                  } else if (length(item) > 1 || is.na(item)) {
                    item[is.na(item)] <- 0
                    return(as.numeric(item[1]))
                  } else {
                    return(as.numeric(item))
                  }
                })))
# Step 4: Add week_num column
mosquito_wide <- wide_data %>%
 mutate(
    week_num = paste0(year(`collection date`), "-", sprintf("%02d", isoweek(`collection date`)))
  )

# Output the transformed data to inspect its structure
str(mosquito_wide)

@@

# Save the wide-format data to the specified path
write_csv(mosquito_wide, file_path_wide)

@@

# Save the wide-format data as an RDS file
saveRDS(mosquito_wide, file_path_wide_rds)

@@

# Read the data from the file
trap_wide_data <- read_csv(trap_wide_path)

trap_wide_data <- trap_wide_data %>%
  mutate(Year = year(as.Date(`collection date`, format = "%Y-%m-%d")))
trap_wide_data
# Assuming all species columns are next to each other, calculate the proportion of MEL
proportion_mel_by_year <- trap_wide_data %>%
  rowwise() %>%
  mutate(Total_All = sum(c_across(ABS:OCH), na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(Year) %>%
  summarise(
    Total_MEL = sum(MEL, na.rm = TRUE),
    Total_All = sum(Total_All, na.rm = TRUE), # Sum of all species by year
    MEL_Prop = Total_MEL / Total_All # Proportion of MEL
  )

@@

# Calculate the proportion of MEL by week
# Assuming all species columns are next to each other, calculate the proportion of MEL by week_num
proportion_mel_by_week <- trap_wide_data %>%
  # Calculate row-wise totals for all species
  rowwise() %>%
  mutate(Total_All = sum(c_across(ABS:OCH), na.rm = TRUE)) %>%
  ungroup() %>%
  # Then group by week_num and summarize
  group_by(week_num) %>%
  summarise(Total_MEL = sum(MEL, na.rm = TRUE),
            Total_All = sum(Total_All, na.rm = TRUE), # Sum of all species by week_num
            MEL_Prop = Total_MEL / Total_All) # Proportion of MEL

@@

```{r}
# Load the datasets
yearly_mel_ave <- readRDS(here("data", "processed-data", "rds", "yearly_mel_average.rds"))
summed_prcp_by_year_filtered <- readRDS(here("data", "processed-data", "rds", "summed_prcp_by_year_filtered_2013_2020.rds"))

@@

plot_mel_precip <- ggplot(data = combined_data) +
  geom_col(aes(x = factor(Year), y = Average_MEL), fill = "blue") +
  geom_line(aes(x = factor(Year), y = Total_PRCP, group = 1), color = "red", linetype = "dashed", linewidth = 1) +
  geom_point(aes(x = factor(Year), y = Total_PRCP), color = "red", size = 3) +
  geom_hline(yintercept = 45.6, linetype = "dotted", color = "darkgreen", size = 1) +
  annotate("text", x = Inf, y = 45.6, label = "Avg Precip: 45.6 inches", 
           hjust = 3.2, vjust = -1.1, color = "darkgreen", size = 3.5, fontface = "bold", angle = 0) +
  scale_y_continuous(name = "Average Count of MEL", 
                     sec.axis = sec_axis(~ ., name = "Total Precipitation (inches)")) +
  labs(title = "Average Count of MEL Mosquitoes and Total Precipitation by Year",
       x = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.title.x = element_text(face = "bold"))

plot_mel_precip

@@

### 2024-04-25 products/manuscript/Manuscript.qmd
further edits and file cleaning

*This project will inform the development of my dissertation prospectus
by identifying significant relationships between variables and guide the
methods development.* Eastern equine encephalitis (EEE) is caused by an
Alphavirus transmitted to humans by the bite of an infected mosquito.
Human infections are rare, but serious. About 30-50% of symptomatic
cases lead to death and those who survive are left with life-long
disability. The transmission cycle for EEE includes two distinct cycles:
an amplifying enzootic cycle in which the virus is transmitted between
the *Culiseta melanura* mosquito and birds and an epizootic cycle in
which bridge vector mosquitoes transmit the EEE virus (EEEv) from birds
to humans and other mammals [@goZoonoticEncephalitidesCaused2014].
Because C. melanura feeds almost exclusively on birds, it is not
considered a direct human threat. However, EEEv isolations in *C.
melanura* are an early warning sign that EEEv is circulating in the
ecosystem. Isolations in bridge vector mosquito species indicate
heightened transmission risk to humans.
:::

@@

Moreover, the influence of other variables remains even less explored.
Birds, especially passerine species, serve as amplifying hosts for EEEV
and WNV, potentially affecting the level of virus present in an area.
Despite their importance in the virus transmission cycle, the effect of
fluctuations in bird populations on the prevalence of these viruses
throughout a season has received minimal attention. Finally, given the
migratory nature of many passerine species, it's worth exploring whether
EEEV activity in the warmer regions of the United States, during the
months preceding mosquito season in Massachusetts, could serve as an
early indicator of virus activity in the state. Understanding these
dynamics is crucial for predicting outbreaks and implementing effective
control measures.

@@

### Human and animal case data

While the goal of this project is to set a foundation for models that
predict human cases of EEE later this year. Human case data will not be
an outcome in the models for this project. However, the case data will
incorporated into the models. Some have hypothesized that EEE activity
in the southern United States could serve as an early warning signal for
EEE activity in Massachusetts. The rationale is that birds that migrate
from the south to Massachusetts could bring the virus with them. In
active EEE years, more infected birds may arrive in Massachusetts and
spread the virus to local mosquitoes at a higher rate. Florida is a
known hotspot for EEE activity and is the only state that reports more
human cases than Massachusetts. The Florida Department of Health reports
EEE activity in animals and humans annually. In addition to the wild and
domestic animal case reports, many mosquito control programs in Florida
maintain flocks of sentinel chickens that are tested for EEE and WNV.
The sentinel chicken data is used to monitor virus activity in the
mosquito population.

### Bird observation data

The North American Breeding Bird Survey (BBS) is conducted annually in
late spring or early summer. The United State Geological Survey (USGS)
maintains the BBS database. The BBS is a long-term, volunteer-driven
bird monitoring program that collects data on bird populations across
North America. The data is used to track the status and trends of bird
populations over time. The BBS data includes information on the number
of birds observed, the location of the observation, and the date of the
observation. The data is collected by trained volunteers who conduct
roadside surveys along predetermined routes. The data is used to
estimate the population size and distribution of bird species across the
continent. The BBS data will be used to estimate the population size and
distribution of bird species in Massachusetts.

Since birds are the amplifying hosts for EEE, their population size and
distribution can have a significant impact on the prevalence of the
virus in the environment. Additionally, the migratory patterns of birds
can influence the spread of the virus across regions. By incorporating
bird observation data into the models, we can better understand the
relationship between bird populations and EEE activity in Massachusetts.

### Temporal scale

For most models in this project, the outcome, MIR, will be calculated by
week. While the main temporal scale for this project is the week, the
data will be aggregated to other temporal scales. The week is the most
common temporal scale for mosquito surveillance data. The week number
will be calculated according to ISO 8601 standards, and stored in a
YYYY-WW format. For example, week 8 in 2019 will be displayed as
2019-08. The week number will provide a common field by which all
datasets with a temporal attribute can be joined. It will also allow for
grouping records and making calculations by week number. This format
helps to standardize temporal scales and make comparisons between years
easier.

While mosquito surveillance and weather data can be aggregated to the
week, other datasets, such as the Annual BBS cannot appropriately be
aggregated by week. Additionally, mosquito surveillance in Massachusetts
is conducted from June to October and EEE activity in Florida occurs
year-round. For these predictors, the temporal scale will be the year.

@@

## Exploratory/Descriptive analysis

It is important to know where the mosquitoes infected with EEE are
captured. Below is a table showing the virus isolations for EEE in
trapped mosquitoes by county between 2014 and 2020.

![](images/eee_isolation_county-04.png){fig-align="center" width="192"}

One factor to consider is if early virus isolations are associated with
cases in humans and other mammals. The table below shows the date of the
first EEE virus isolation in mosquitoes and the number of human of other
animal cases by year.

@@

Unfortunately, the time horizon is limited by data availability.
Additionally, EEE human infections are rare, so traditional statistical
methods may be inadequate to suggest any relationships. What is clear is
that 2019 was an exceptional year for EEE.

@@

## Basic statistical analysis

The initial tests have been basic regression models.

The first models explored the relationship between virus isolation in
mosquitoes and weather conditions (average temperature by week number
and total precipitation by week). The outcome was whether there was at
least 1 positive test for pool in a given week. The predictor in the
first model was the average maximum daily temperature the week before
the mosquito collection week. The logistic model suggests that there was
a relationship between the two. (The R code for the model can be found
in the analysis-code file: analysis.qmd) The average temperature in the
week before is positively associated with the likelihood of a positive
result, with an odds ratio of approximately 1.069 for each unit increase
in temperature (Estimate = 0.06695, SE = 0.02033, p \< 0.001),
indicating that higher temperatures significantly increase the
probability of a positive outcome. The model's improvement is
highlighted by a decrease in residual deviance to 405.79 from the null
deviance of 417.24, upon incorporating Lagged_Avg_TMAX as a predictor,
demonstrating a better fit to the observed data. However, with a
residual deviance of 405.79 on 299 degrees of freedom, it is evident
that not all variability in the outcome is explained, suggesting the
presence of other influential factors. The model's AIC of 411.59
indicates a balance between model complexity and predictive power,
supporting its relative utility in understanding the data. However, this
relationship should be examined more closely as temperature is likely
auto correlated with progressing weeks during summer months. While
precipitation has been demonstrated to influence the population dynamics
it did not show to have any effect on a positive result. This was
initailly surprising, but upon after considering the specific habitat of
the main EEE enzootic vector, *Culiseta melanura*, it makes sense. MEL
larvae develop in well insulated crypts that are less sensitive
evaporative processes than surface water. The effect of precipitation on
MEL larval habitat abundance may be less pronounced than other vector
species that develop in open, standing water, like ditches and
containers. Another model assessed whether the temperature precipitation
in April, May, and June of each year effected the date of the first
positive pool test (R file: PT3_edacode). The date of the first positive
pool was identified and assigned a day number from January 1st of that
year. The predictors in this linear model were the average maximum
temperature for April, May, and June and the sum of precipitation for
the same time period. None of these predictors were significant.

>>>>

Eastern equine encephalitis (EEE) is a rare but serious disease caused by an Alphavirus transmitted to humans by infected mosquitoes. With a fatality rate of 30-50% and potential lifelong disability for survivors, understanding its transmission cycle is crucial. EEE transmission involves two cycles: an enzootic cycle between Culiseta melanura mosquitoes and birds, and an epizootic cycle where bridge vector mosquitoes transmit the virus to humans and other mammals. While C. melanura poses minimal direct threat to humans, isolations of EEE virus in these mosquitoes serve as early indicators of virus circulation. Conversely, isolations in bridge vector species signal heightened transmission risk. This research aims to identify significant relationships between variables to inform dissertation prospectus development and guide methodological approaches.
:::

@@

@@

@@

## Exploratory/Descriptive analysis
One factor to consider is if early virus isolations are associated with
cases in humans and other mammals. The table below shows the date of the
first EEE virus isolation in mosquitoes and the number of human of other
animal cases by year.

@@

@@

## Basic statistical analysis
Basic linear models assessed environmental factors' impact on the EEE virus transmission cycle. Initial models investigated the link between virus isolation in mosquitoes and weather conditions (average temperature by week and total precipitation). The outcome was whether at least one positive test occurred for a pool in a given week. The predictor in the first model was the average maximum daily temperature from the previous week. The logistic model revealed a significant relationship. Higher temperatures were positively associated with the likelihood of a positive result (odds ratio ≈ 1.069 for each unit increase in temperature; Estimate = 0.06695, SE = 0.02033, p < 0.001). Incorporating Lagged_Avg_TMAX as a predictor improved the model, reducing residual deviance to 405.79 from the null deviance of 417.24, indicating a better fit. However, residual deviance on 299 degrees of freedom suggests unexplained variability, indicating other influential factors. The model's AIC of 411.59 suggests a balance between complexity and predictive power, supporting its utility. Yet, this relationship warrants closer examination due to likely autocorrelation of temperature with progressing weeks during summer months

While
precipitation has been demonstrated to influence the population dynamics
it did not show to have any effect on a positive result. This was
initailly surprising, but upon after considering the specific habitat of
the main EEE enzootic vector, *Culiseta melanura*, it makes sense. MEL
larvae develop in well insulated crypts that are less sensitive
evaporative processes than surface water. The effect of precipitation on
MEL larval habitat abundance may be less pronounced than other vector
species that develop in open, standing water, like ditches and
containers. Another model assessed whether the temperature precipitation
in April, May, and June of each year effected the date of the first
positive pool test (R file: PT3_edacode). The date of the first positive
pool was identified and assigned a day number from January 1st of that
year. The predictors in this linear model were the average maximum
temperature for April, May, and June and the sum of precipitation for
the same time period. None of these predictors were significant.
